The Magnetorotational Instability Prefers Three Dimensions

### 2019-07-22 paper/paper.tex
Fix long equation.
Right asymptotic equation this time...
More words.

\section{Introduction}
\label{sec:intro}
The magnetorotational instability (MRI) is an extremely important instability in astrophysical fluid dynamics.
By changing the stabiliity criterion for differentially rotating flows from a negative angular \emph{momentum} gradient to a negative angular \emph{velocity} gradient, a weak magnetic field allows the Keplerian velocity profile due to a central point mass to drive turbulence \citep[e.g.][]{1998RvMP...70....1B}.
This discovery provided a pathway to explain the ubiquitous accretion onto compact objects at rates compatible with observations and may play a significant role in the formation of planets \citep[e.g.][]{2007Natur.448.1022J}.
In these objects, the gravitational field dominates the local plasma dynamics, and thus the MRI cannot significantly affect the background shear.
It must saturate by other means in these situations \citep{2018MNRAS.474.3451X}.
However, stars have differential rotation profiles driven by much weaker stresses.
In these regions, the MRI could be active and because the driving is much weaker, it can sature by pushing the background shear close to critical \citep{2017ApJ...841....1C,2017ApJ...841....2C}.

@@

\section{Method}
\label{sec:method}
We solve the linearized magnetohydrodynamic equations in a frame rotating with angular frequency $\Omega$. The Navier-Stokes equation is in a fairly standard form,
\begin{equation}
  \label{eq:ns}
  \frac{D \mathbf{v}}{Dt} - f \hat{z} \times \mathbf{v} + S v_x \hat{z} + \mathbf{\nabla}{p} + \nu \mathbf{\nabla} \times \mathbf{\omega} = 0,
\end{equation}
where $\omega = \mathbf{\nabla} \times \mathbf{v}$ is the vorticity, $f = 2 \Omega$ is the Coriolis parameter, and $S$ is the background shear rate. $Df/Dt = \partial f/\partial t + S x \partial f/\partial y$ is the advective derivative including the background linear shear velocity.  We write the induction equation in terms of the $x$ component of the magnetic field,
\begin{equation}
  \label{eq:Bx}
  \frac{D B_x}{Dt} - B_0 \partial_z v_x + \eta (\partial_y J_z - \partial_z J_y) = 0
\end{equation}
and the $x$ component of the current density,
\begin{equation}
  \label{eq:Jx}
  \frac{D J_x}{Dt} - B_0 \partial_z \omega_x + S \partial_z B_x - \eta \nabla^2 J_x = 0
\end{equation}
We explicitly enforce the divergence constraint on the incompressible velocity field $v$ and the magnetic field,
\begin{equation}
  \label{eq:divu}
  \mathbf{\nabla} \cdot \mathbf{v} = \mathbf{\nabla} \cdot \mathbf{B} = 0.
\end{equation}
The magnetic field is expressed in Alfven units, $\mathbf{B} = \mathbf{B_*}/\sqrt{\mu_0 \rho}$.
We consider the behavior of the MRI in a vertically ($z$) and horizontally ($y$) periodic channel of width $L_x = \pi$. The walls at $x = \pm L_x/2$ are perfectly conducting and stress-free. The MRI is a weak-field instability; in the inviscid, ideal case the critical shear rate for instability in two dimensions is given by
\begin{equation}
  \label{eq:Sc}
  S_c = \frac{-\pi^2 B^2}{f d^2}.
\end{equation}
Here, we use $\SSC$ as our control parameter.

We solve equations~(\ref{eq:ns})--~(\ref{eq:divu}) by assuming harmonic perturbations in $y$ and $z$, $f = \hat{f}(x) e^{i(k_y y + k_z z) + \gamma t}$. This reduces to an eigenvalue problem in $x$, which we solve using the \emph{Dedalus} framework. The advantage of this particular form is that it can be expressed as a set of ten first-order equations with Dirichlet boundary conditions for the stress-free, perfect conductor boundary conditions we use.

Our interest is in nearly ideal ($\eta = 0$), inviscid ($\nu = 0$) conditions.
However, in order to avoid critical layers in the stable solutions, we set $\eta=\nu=10^{-5}$.
We have confirmed our main results are insensitive to this particular value.

\section{Results}
\label{sec:results}
Our most surprising result is that at very small values of supercriticality, the fastest growing mode is not axisymmetric. 
Figure~\ref{fig:growth_rate}a shows the growth rates for four values of $\SSC$. 
At $\SSC = 1.02$, the maximum growth rate is $(k_y, k_z) \simeq (0.263, 0.447)$.
Figure~\ref{fig:growth_rate}b compares the numerically growth rates for $\SSC=1.002$ between $0 <= k_y < 0.2$ to an asymptotic calculation.
By expanding equations~(\ref{eq:ns}-~\ref{eq:divu}) in terms of $S = \SSC + \epsilon R$, we compute the leading order correction to the growth rate from three-dimensional effects as
\begin{equation}
  \label{eq:asymp}
  \left(\pi ^2 B^2+d^2 f^2 \right) \omega^{2} - B^4 k^2 \left(d^2 k^2-\pi ^2 R\right)   \ + \ \frac{\pi ^2 B^4 \ell ^2}{12}  \left(\left(6+\pi ^2\right) q+\pi ^2-6\right) \ = \ 0.
\end{equation}

%
\onecolumngrid

\begin{figure*}[b!]
  \includegraphics[width=\textwidth]{fig_1.pdf}
  \caption{Stuff}
  \label{fig:growth_rate}
\end{figure*}
\twocolumngrid
%
Numerical simulations have consistently shown axisymmetric modes dominating early evolution of the MRI before parasitic modes cause a breakdown into MHD turbulence \citep[e.g.][]{1995ApJ...440..742H,2018ApJ...853..174H,2019ApJS..241...26D}. 
In order to reconcile our results with these simulations, in figure~\ref{fig:alpha} we plot the angle $\alpha = \arctan k_y/k_z$ of the fastest growing mode as function of $\SSC$.
Above $\SSC \gtrsim 2.05$, $\alpha$ is zero, indicating that axisymmetric modes have the fastest growth rates.
For the fiducial run in \citet{1996ApJ...464..690H}, $\SSC \simeq 4.84$ (similar values are found in most works); thus we recover these results.

\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{alpha_vs_ssc_grid.pdf}
  \caption{Angle $\alpha$ vs $\SSC$. Stability limit is at $\SSC \simeq 0.102$}
  \label{fig:alpha}
\end{figure}

\section{Non-Normality}
\label{sec:non-normality}

\section{Speculations}
\label{sec:speculations}

\begin{itemize}
\item non-normal and $k_y$ even when $\alpha = 0$ implies important role for linear dynamics at early times. 
\item Following Squire + Bhattacharjee, this suggests that important stellar layers may in fact be driven by linear dynamics
\item why is MRI non-oscillatory when hydro shear is oscil or stable?
\end{itemize}
The linear operator for the MRI, like other shearing systems, is non-normal. 
%If in two-column mode, this environment will change to single-column
% format so that long equations can be displayed. Use
% sparingly.
%\begin{widetext}
% put long equation here
%\end{widetext}

>>>>

The magnetorotational instability (MRI) is an extremely important instability in astrophysical fluid dynamics.
By changing the stabiliity criterion for differentially rotating flows from a negative angular \emph{momentum} gradient to a negative angular \emph{velocity} gradient, a weak magnetic field allows the Keplerian velocity profile due to a central point mass to drive turbulence \citep[e.g.][]{1998RvMP...70....1B}.
This discovery provided a pathway to explain the ubiquitous accretion onto compact objects at rates compatible with observations and may play a significant role in the formation of planets \citep[e.g.][]{2007Natur.448.1022J}.
In these objects, the gravitational field dominates the local plasma dynamics, and thus the MRI cannot significantly affect the background shear.
It must saturate by other means in these situations \citep{2018MNRAS.474.3451X}.
However, stars have differential rotation profiles driven by much weaker stresses.
In these regions, the MRI could be active and because the driving is much weaker, it can sature by pushing the background shear close to critical \citep{2017ApJ...841....1C,2017ApJ...841....2C}.

@@

We solve the linearized magnetohydrodynamic equations in a frame rotating with angular frequency $\Omega$. The Navier-Stokes equation is in a fairly standard form,
\begin{equation}
  \label{eq:ns}
  \frac{D \mathbf{v}}{Dt} - f \hat{z} \times \mathbf{v} + S v_x \hat{z} + \mathbf{\nabla}{p} + \nu \mathbf{\nabla} \times \mathbf{\omega} = 0,
\end{equation}
where $\omega = \mathbf{\nabla} \times \mathbf{v}$ is the vorticity, $f = 2 \Omega$ is the Coriolis parameter, and $S$ is the background shear rate.
It is common in the accretion disk literature to relate $S$ and $f$ via the $q = S/f$ parameter; for all runs here, $q = 0.75$.
$Df/Dt = \partial f/\partial t + S x \partial f/\partial y$ is the advective derivative including the background linear shear velocity.  We write the induction equation in terms of the $x$ component of the magnetic field,
\begin{equation}
  \label{eq:Bx}
  \frac{D B_x}{Dt} - B_0 \partial_z v_x + \eta (\partial_y J_z - \partial_z J_y) = 0
\end{equation}
and the $x$ component of the current density,
\begin{equation}
  \label{eq:Jx}
  \frac{D J_x}{Dt} - B_0 \partial_z \omega_x + S \partial_z B_x - \eta \nabla^2 J_x = 0
\end{equation}
We explicitly enforce the divergence constraint on the incompressible velocity field $v$ and the magnetic field,
\begin{equation}
  \label{eq:divu}
  \mathbf{\nabla} \cdot \mathbf{v} = \mathbf{\nabla} \cdot \mathbf{B} = 0.
\end{equation}
The magnetic field is expressed in Alfven units, $\mathbf{B} = \mathbf{B_*}/\sqrt{\mu_0 \rho}$.
We consider the behavior of the MRI in a vertically ($z$) and horizontally ($y$) periodic channel of width $L_x = \pi$. The walls at $x = \pm L_x/2$ are perfectly conducting and stress-free. The MRI is a weak-field instability; in the inviscid, ideal case the critical shear rate for instability in two dimensions is given by
\begin{equation}
  \label{eq:Sc}
  S_c = \frac{-\pi^2 B^2}{f d^2}.
\end{equation}
Here, we use $\SSC$ as our control parameter.

We solve equations~(\ref{eq:ns})--~(\ref{eq:divu}) by assuming harmonic perturbations in $y$ and $z$, $f = \hat{f}(x) e^{i(k_y y + k_z z) + \sigma t}$, where $\sigma = \gamma + i\omega$ with $\gamma$ and $\omega$ real. This reduces to an eigenvalue problem in $x$, which we solve using the \emph{Dedalus} framework. The advantage of this particular form is that it can be expressed as a set of ten first-order equations with Dirichlet boundary conditions for the stress-free, perfect conductor boundary conditions we use. 
Our interest is in nearly ideal ($\eta = 0$), inviscid ($\nu = 0$) conditions.
However, in order to avoid critical layers in the stable solutions, we set $\eta=\nu=10^{-5}$.
We have confirmed our main results are insensitive to this particular value.

Our most surprising result is that at small values of supercriticality, the fastest growing mode is not axisymmetric. 
Figure~\ref{fig:growth_rate}a shows the growth rates for four values of $\SSC$. 
At $\SSC = 1.02$, the maximum growth rate is found at $(k_y, k_z) \simeq (0.263, 0.447)$.
By expanding equations~(\ref{eq:ns}-~\ref{eq:divu}) in terms of $S = \SSC + \epsilon R$, we compute the leading order correction to the growth rate from three-dimensional effects as
\begin{widetext}
\begin{equation}
  \label{eq:asymp}
\omega^{2} \ = \ \frac{q\,B^{2}}{(q+1)} \left[ k_z^2 \left(\frac{d^2}{\pi^{2}} k_z^2-R\right) \ - \ \frac{\left(6+\pi ^2\right) q+\pi
   ^2-6}{12 } \, k_{y}^{2}  \right] \ + \ ...
\end{equation}
\end{widetext}
Figure~\ref{fig:growth_rate}b compares the numerically growth rates for $\SSC=1.002$ between $0 <= k_y < 0.2$ to this asymptotic approximation, showing that the numerical calculations accurately reproduce the analytic ones in the region where the latter is valid.
All unstable modes are purely growing (figure~\ref{fig:growth_rate}c); the stable modes are rotationally modified Alfven waves found in left- and right- going pairs, consistent with the analytic predictions for two-dimensional stability calculations.
%
\onecolumngrid

\begin{figure*}[b!]
  \includegraphics[width=\textwidth]{fig_1.pdf}
  \caption{Stuff}
  \label{fig:growth_rate}
\end{figure*}
\twocolumngrid
%
Numerical simulations have consistently shown axisymmetric modes dominating early evolution of the MRI before parasitic modes cause a breakdown into MHD turbulence \citep{1995ApJ...440..742H,2018ApJ...853..174H,2019ApJS..241...26D}. 
In order to reconcile our results with these simulations, in figure~\ref{fig:alpha} we plot the angle $\alpha = \arctan k_y/k_z$ of the fastest growing mode as function of $\SSC$.
Above $\SSC \gtrsim 2.05$, $\alpha$ is zero, indicating that axisymmetric modes have the fastest growth rates.
For the fiducial run in \citet{1996ApJ...464..690H}, $\SSC \simeq 4.84$ (similar values are found in most works); thus we recover these results.
%
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{alpha_vs_ssc_grid.pdf}
  \caption{Angle $\alpha$ vs $\SSC$. Stability limit is at $\SSC \simeq 0.102$}
  \label{fig:alpha}
\end{figure}
%
However, even for $\SSC> 2.05$ there are significant swaths of unstable three-dimensional modes with growth rates comparable to the two-dimensional mode (figure~\ref{fig:growth_rate}a).

The linear operator for the MRI, like other shearing systems, is non-normal.
As such, significant non-modal transient amplification occurs even for stable systems.

% \begin{itemize}
% \item non-normal and $k_y$ even when $\alpha = 0$ implies important role for linear dynamics at early times. 
% \item Following Squire + Bhattacharjee, this suggests that important stellar layers may in fact be driven by linear dynamics
% \item why is MRI non-oscillatory when hydro shear is oscil or stable?
% \end{itemize}

%If in two-column mode, this environment will change to single-column
% format so that long equations can be displayed. Use
% sparingly.
%\begin{widetext}
% put long equation here
%\end{widetext}
